Global Warming Distributed Data Analysis

### 2016-03-28 report/article.tex
corrections

\subsection{Temperature data quality}
\label{sec:filter}
Before using the actual temperature data, we filtered it through a custom quality check. Each measurement come with a quality flag set at a different value depending on how reliable the measurement was\cite{noaaDataFormat}.
It is important to use this flag to remove wrong data in the dataset, as reported in the a document provided by NOAA that describes the problem inside the dataset \cite{noaaProblems}: ``Various, mostly random errors, are present in the overall dataset, and these are being documented and scheduled for correction. Quality control of the data
(which has already taken place and reflected in the online data) has corrected
well over 99 percent of the errors present in the original data prior to the quality control.''
In particular we accepted all the measurement with the following quality code value:
\begin{itemize}
    \item 0: Passed gross limits check;
    \item 1: Passed all quality control checks;
    \item 4: Passed gross limits check, data originate from an NCEI data source;
    \item 5: Passed all quality control checks, data originate from an NCEI data source;
    \item A: Data value flagged as suspect, but accepted as a good value;
    \item I: Data value not originally in data, but inserted by validator;
    \item M: Manual changes made to value based on information provided by NWS or FAA;
    \item P: Data value not originally flagged as suspect, but replaced by validator;
    \item R: Data value replaced with value computed by NCEI software;
    \item U: Data value replaced with edited value.
\end{itemize}
We opted for a loose policy on the data quality, accepting also values that didn't pass all the quality checks or that were manually or software replaced. We refuted all the codes that stand for erroneous, missing or suspect data that was not manually or automatically checked and replaced by the institution. Furthermore, we applied a bound check for all the values greater than the maximum registered temperature (+61.8 \celsius) and the minimum one (-93.2 \celsius). \\
The data collected can be of two different types depending on the origin: from a fixed weather station or from a mobile one. In the second case, it's possible that the data registered from a mobile station (e.g. on a ship) is reported at a certain position and the following measurement is reported at a different position. Since we didn't distinguish data by time of the day, it's important that the number of measurements from a certain position is approximately the same for different hours. Since this is not guaranteed by moving stations, we decided to filter the measurement obtained from this source. \\

@@

The data stored in the distributed file system is grouped in folders by year. As shown in figure \ref{fig:dataSize} the size of the data to analyze is very different between the years. Inside each folder, there is a file for each station that reported measurements. HDFS is not very efficient when working with many small files\cite{zhang2012improving}, as in our case. The solution we opted to use is to merge the small files in bigger files that have the same size as Hadoop's default block size. To merge the files we used filecrush\cite{filecrush}. This tool merges small files, located inside an HDFS instance, into bigger files with the size of a multiplier of an HDFS block. We have run this tool one time per each folder of the data. The number of total files was reduced by an order of magnitude of $10^2$.\\

@@

The chart in figure \ref{fig:stationsRegion} shows the surface coverage of the selected regions for each year from the data we analyzed. As we can see, we don't have data from every part of the regions, with Europe and US the more covered ones (with peaks of 70\% and 60\%). The are two main reason behind this:
\begin{itemize}
    \item In the chart we took into account the number of regions for which we have data on the total amount of regions in the area. With a different selection of region size we could increase or decrease the total coverage (at the expense of precision).
    \item There may be areas in the regions where it's more difficult to have an uniform stations coverage because of geomorphological reasons. For instance, there are far less regions covered in the middle part of Australia due to its desert climate, that makes it more difficult to install, maintain and monitor the stations.
\end{itemize}

@@

In the graphic interface there is the possibility to select the two set of five-year results to compare. The heatmap shows a colored square in a latitude and longitude quadrant if it's present a data for that quadrant in both the two sets of results. Till the period 1921-1925 there are few data in the results, with consequently few colored squares in the map. In the interface there is also the possibility to change the temperature metric that is used to compared the two set of results.\\
One problem of the interface is that the colored squares used in the map should change their height depending on the latitude value where they are positioned because the map projection isn't equiareal. The consequence in the heatmap visualization is that there is some empty space between the squares of a certain latitude value and the squares with the value of the latitude increased or decreased by one. This aspect makes the heatmap looks like as it is missing some data for certain latitude when instead the data are present, just the squares should be rectangles with a bigger height value.\\

@@

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{increaseEurope}
\caption{Map that shows the increase of the average temperature in Europe and in Asia between 1976 and 2015.}
\label{fig:interface}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{increaseGlobal}
\caption{Map that shows the increase of the minimum temperature globally between 1976 and 2015.}
\label{fig:interface}
\end{figure}

@@

\subsubsection{Standard Deviation analysis}
\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stdev}
\caption{Standard deviation for Average, Minimum and Maximum values per region.}
\label{fig:stdev}
\end{figure}
In figure \ref{fig:stdev} we plotted the value of the standard deviation per statistics and per region. It's clear that the values of maximum and minimum have larger fluctuation than the average. This may be caused by one or a few erroneous data report caused by environmental accidents. Interesting to note that the minimum value have a way larger standard deviation that the maximum. By correlating this with the stations distribution we came to the conclusion that this is caused by the installation of new stations in more remote and difficult to access regions (e.g. Arctic ocean and mountain regions). \\
Moreover, the minimum standard deviation for the Australian region was suspect, because the lowest temperature registered for the region is -23.4\degree C\cite{ausTemp}, so it's difficult to expect a standard deviation similar to that of the European region or even higher that that of the American one. After analyzing the data we found out erroneous data coming from the south east region of Australia reporting temperature of around -70\degree C.

@@

For what concerns the metrics used for analyzing the data, the average temperature is the most useful metric. It shows clearly the increase of the temperature in different periods and it's stable also if some particular events happen in short periods of the year. Minimum and maximum temperature are more susceptible to those events and indeed they show less clearly the increase of the temperature in the periods where the average metric reveals the increase. The standard deviation metric is useful to understand if there are errors in the data or if the data are not uniformly distributed (e.g. when the distributions of the stations is changed during the five year of the grouped results then the standard deviation results higher).

@@

\includegraphics[width=1\linewidth]{stations1907}
\caption{Wheater stations used in 1907. In the first years we only have measurements coming from stations in Finland}
\label{fig:stations1907}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1930}
\caption{Wheater stations used in 1930. Here we can see the stations cover the most part of central Europe. The problem we encountered is that there is no continuity in the stations that are used between consecutive years, so it's difficult to make comparisons.}
\label{fig:stations1930}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1945}
\caption{Wheater stations used in 1945. Around this year we start to have a relevant number of stations that cover most part of the surface.}
\label{fig:stations1945}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1968}
\caption{Wheater stations used in 1968. In the late 60s we notice a sudden drop in the earth coverage compared to the previous years. We took this into account when making the analysis.}
\label{fig:stations1968}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1972}
\caption{Wheater stations used in 1972. As shown in figure \ref{fig:stations}, we noticed a sudden drop in the number of stations for this year. In particular we observe the absence of stations from east Europe and the ex Soviet Union Area.}
\label{fig:stations1972}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1980}
\caption{Wheater stations used in 1980. From this year onwards we have a large number of stations covering most part of the surface.}
\label{fig:stations1980}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations2014}
\caption{Wheater stations used in 2014.}
\label{fig:stations2014}
\end{figure}

>>>>

\subsection{Temperature data quality}
\label{sec:filter}
Before using the actual temperature data, we filtered it through a custom quality check. Each measurement come with a quality flag set at a different value depending on how reliable the measurement was\cite{noaaDataFormat}.
It is important to use this flag to remove wrong data in the dataset, as reported in the document provided by NOAA that describes the problem inside the dataset \cite{noaaProblems}: ``Various, mostly random errors, are present in the overall dataset, and these are being documented and scheduled for correction. Quality control of the data
(which has already taken place and reflected in the online data) has corrected
well over 99 percent of the errors present in the original data prior to the quality control.''
In particular we accepted all the measurement with the following quality code value:
\begin{itemize}
    \item 0: Passed gross limits check;
    \item 1: Passed all quality control checks;
    \item 4: Passed gross limits check, data originate from an NCEI data source;
    \item 5: Passed all quality control checks, data originate from an NCEI data source;
    \item A: Data value flagged as suspect, but accepted as a good value;
    \item I: Data value not originally in data, but inserted by validator;
    \item M: Manual changes made to value based on information provided by NWS or FAA;
    \item P: Data value not originally flagged as suspect, but replaced by validator;
    \item R: Data value replaced with value computed by NCEI software;
    \item U: Data value replaced with edited value.
\end{itemize}
We opted for a loose policy on the data quality, accepting also values that didn't pass all the quality checks or that were manually or software replaced. We refuted all the codes that stand for erroneous, missing or suspect data that was not manually or automatically checked and replaced by the institution. Furthermore, we applied a bound check for all the values greater than the maximum registered temperature (+61.8 \celsius) and the minimum one (-93.2 \celsius). \\
The data collected can be of two different types depending on the origin: from a fixed weather station or from a mobile one. In the second case, it's possible that the data registered from a mobile station (e.g. on a ship) is reported at a certain position and the following measurement is reported at a different position. Since we didn't distinguish data by time of the day, it's important that the number of measurements from a certain position is approximately the same for different hours. Since this is not guaranteed by moving stations, we decided to filter the measurement obtained from this source. \\

@@

The data stored in the distributed file system is grouped in folders by year. As shown in figure \ref{fig:dataSize} the size of the data to analyze is very different between the years. Inside each folder, there is a file for each station that reported measurements. HDFS is not very efficient when working with many small files\cite{zhang2012improving}, as in our case. The solution we opted to use is to merge the small files in bigger files that have the same size as Hadoop's default block size. To merge the files we used filecrush\cite{filecrush}. This tool merges small files, located inside an HDFS instance, into bigger files with the size of a multiplier of an HDFS block. We have run this tool one time for each folder of the data. The number of total files was reduced by an order of magnitude of $10^2$.\\

@@

The chart in figure \ref{fig:stationsRegion} shows the surface coverage of the selected regions for each year from the data we analyzed. As we can see, we don't have data from every part of the regions, with Europe and US the more covered ones (with peaks of 70\% and 60\%). The are two main reason behind this:
\begin{itemize}
    \item In the chart we took into account the number of regions for which we have data on the total amount of regions in the area. With a different selection of region size we could increase or decrease the total coverage (at the expense of precision).
    \item There may be areas in the regions where it's more difficult to have a uniform stations coverage because of geomorphological reasons. For instance, there are far less regions covered in the middle part of Australia due to its desert climate, which makes it more difficult to install, maintain and monitor the stations.
\end{itemize}

@@

In the graphic interface there is the possibility to select the two set of five-year results to compare. The heatmap shows a colored square in a latitude and longitude quadrant if it's present a data for that quadrant in both the two sets of results. Till the period 1921-1925 there are few data in the results, with consequently few colored squares in the map. In the interface there is also the possibility to change the temperature metric that is used to compare the two set of results.\\
One problem of the interface is that the colored squares used in the map should change their height depending on the latitude value where they are positioned because the map projection isn't equiareal. The consequence in the heatmap visualization is that there is some empty space between the squares of a certain latitude value and the squares with the value of the latitude increased or decreased by one. This aspect makes the heatmap looks like as it is missing some data for certain latitude when instead the data are present, just the squares should be rectangles with a bigger height value.\\

@@

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{increaseEurope}
\caption{Map that shows the increase of the average temperature in Europe and in Asia between 1976 and 2015.}
\label{img:increaseMap}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{increaseGlobal}
\caption{Map that shows the increase of the minimum temperature globally between 1976 and 2015.}
\label{img:increaseGlobal}
\end{figure}

@@

\subsubsection{Standard Deviation analysis}
\begin{figure}[tbh]
\label{fig:stdev}
\includegraphics[width=1\linewidth]{stdev}
\caption{Standard deviation for Average, Minimum and Maximum values per region.}
\end{figure}
In figure \ref{fig:stdev} we plotted the value of the standard deviation per statistics and per region. It's clear that the values of maximum and minimum have larger fluctuation than the average. This may be caused by one or a few erroneous data report caused by environmental accidents. Interesting to note that the minimum value have a way larger standard deviation that the maximum. By correlating this with the stations distribution we came to the conclusion that this is caused by the installation of new stations in more remote and difficult to access regions (e.g. Arctic ocean and mountain regions). \\
Moreover, the minimum standard deviation for the Australian region was suspect, because the lowest temperature registered for the region is -23.4\degree C\cite{ausTemp}, so it's difficult to expect a standard deviation similar to that of the European region or even higher than that of the American one. After analyzing the data we found out erroneous data coming from the south east region of Australia reporting temperature of around -70\degree C.

@@

For what concerns the metrics used for analyzing the data, the average temperature is the most useful metric. It shows clearly the increase of the temperature in different periods. Also, it's stable even if some particular events happen in short periods of the year. Minimum and maximum temperature are more susceptible to those events and indeed they show less clearly the increase of the temperature in the periods where the average metric reveals the increase. The standard deviation metric is useful to understand if there are errors in the data or if the data are not uniformly distributed (e.g. when the distributions of the stations is changed during the five year of the grouped results then the standard deviation results higher).

@@

\includegraphics[width=1\linewidth]{stations1907}
\caption{Weather stations used in 1907. In the first years we only have measurements coming from stations in Finland}
\label{fig:stations1907}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1930}
\caption{Weather stations used in 1930. Here we can see the stations cover the most part of central Europe. The problem we encountered is that there is no continuity in the stations that are used between consecutive years, so it's difficult to make comparisons.}
\label{fig:stations1930}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1945}
\caption{Weather stations used in 1945. Around this year we start to have a relevant number of stations that cover most part of the surface.}
\label{fig:stations1945}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1968}
\caption{Weather stations used in 1968. In the late 60s we notice a sudden drop in the earth coverage compared to the previous years. We took this into account when making the analysis.}
\label{fig:stations1968}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1972}
\caption{Weather stations used in 1972. As shown in figure \ref{fig:stations}, we noticed a sudden drop in the number of stations for this year. In particular we observe the absence of stations from east Europe and the ex Soviet Union Area.}
\label{fig:stations1972}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations1980}
\caption{Weather stations used in 1980. From this year onwards we have a large number of stations covering most part of the surface.}
\label{fig:stations1980}
\end{figure}

\begin{figure}[tbh]
\includegraphics[width=1\linewidth]{stations2014}
\caption{Weather stations used in 2014.}
\label{fig:stations2014}
\end{figure}
